Convergent Evolution During Local Adaptation to Patchy Landscapes

### 2015-08-30 patchy-review-responses-second-round.tex

\vskip 4em

\reply

The main change,
which addresses most of the concerns,
is that we uncovered an error in the derivation of equation \eqref{eqn:migrate},
that supplied the missing factor of area noted by the reviewers.
The error occurred in the transition between section \nameref{ss:heuristics} 
and section \nameref{ss:hitting_occupation}
because in the former we had tried to keep things simple
by not including shape of the patch in
gestalt equations \eqref{eqn:gestalt_q} and \eqref{eqn:gestalt_migrate},
but then (more correctly) treated the shape in the latter;
effectively we were treating the patch as very small in one place
and large in another.
We've rewritten the argument in \nameref{ss:heuristics} 
to deal explicitly with the integral over the whole patch,
which also necessitated adding a new (short) appendix \nameref{apx:qS}.
Section \nameref{ss:hitting_occupation} required very little change.

We also turned up a minor bug in the discussion of conditioned branching processes:
the exact picture is not quite so simple,
but the approximate picture is still the same --
see around \llname{ll:branching_process_fix}.

@@

%%%%%%
\point{}{
Regarding the simulations presented in Fig 4, I don't think they strongly support the functional form of $\migrate$. Points with very similar observed migration time have predicted migration times that differ by a factor of a 100 (the more extreme are outside the validity of the approximations, but still I don't think this figure is strong evidence that the math is correct.)
}

\reply
We think the main issue here was inclusion of points outside the validity of the approximations,
since of the parameters falling in the range of validity discussed in \nameref{ss:assumptions},
the largest discrepancy was a factor of 7, not 100.
We've re-made this figure only using parameter combinations that fall within the range of validity
($R > \sigma/\sqrt{s_m}$, patch width larger than $\sigma/\sqrt{s_m} \atan(\sqrt{s_m/s_b})$, and $2 s_m \sigma N > 1$);
%         ( R*sqrt(abs(sm))/sigma > 1 ) & 
%         ( size2 > (sigma/sqrt(abs(sm)))*atan(sqrt(abs(sm/gb))) ) &
%         ( abs(sm) * 2 * sigma * N > 1 )
the figures that include all simulation results are still included in the supplement
(and hopefully will inspire theory that is applicable in the other regions).

%%%%%%
\point{eq.~56 is wrong}{
    as easily seen with $\Theta=0$ \ldots
    so eq.~56 has an extra 2, resulting from a wrong expansion of $\sqrt{1+x}$,
    and so eq.~57--61 are also wrong.
}

\reply
Thanks for catching that mistake.  
We removed the extra factor of 2 (section~\secref{apx:outflux}).

@@

%%%%%%
\point{time in transit}{
    One can obtain these results by a simple saddle point approximation:
    \begin{align*}
        \P\{ \text{getting to $x$ in time $t$} \} &= \P\{\text{diffusion to $x$ in time $t$}\} \P\{\text{surviving}\} \\
                                                  &= \frac{1}{(2\pi\sigma^2)^{d/2}} e^{-\frac{x^2}{2\sigma^2 t}-st}
    \end{align*}
    the exponent is minimal when
    \[
        \frac{x^2}{2\sigma^2 t^2} = s \quad \Rightarrow \quad t^* = \frac{x}{\sigma\sqrt{2s}},
    \]
    this gives your eq.~17.
    The variance is simply the inverse of the second derivative of the exponent
    \[
        \frac{1}{\var[\tau]} = \frac{x^2}{\sigma^2 (t^*)^3} = \frac{\sigma}{x(2s)^{3/2}} .
    \]
}

\reply
This is a nice point;
perhaps which approach should be included
comes down to personal taste
as to whether ``saddle point approximation'' or ``Laplace transform'' is a simpler idea?  
We decided to stick with the Laplace transform.

%%%%%%
\point{now to the thornier issues of dimension, geometry, and fluxes}{
    \begin{itemize}
        \item {} [with $\ell = \sigma/\sqrt{2s}$, $r$ the radius of the patch, $R$ the distance between patches, and $\ell \ll r \ll R$]
        \item the average density is $q(x) \sim e^{-x\sqrt{2s}/\sigma} = e^{-x/\ell}$
        \item but fluctuates far from the source as [families of size $K$, with local density $k/\ell^d$]
        \item families of size $1/s$ live for $1/s$ generations and hence contribute $1/s^2$ to the time-averaged density $q(x)$
        \item \[ q(x) = \lambda_\text{arr}(x) / s^2 \ell^d, \]
            where $\lambda_\text{arr}(x)$ is the rate at which families arrive that overlap with $x$
        \item \[ \lambda_\text{arr}(x) = q(x) s^2 \ell^d \]
            and
            \[ \Lambda_\text{arr}(x) = \lambda(x) \ell^d = q(x) s^2 \]
            would be an ``arrival density''
        \item to get at $\migrate$, arrival needs to be integrated over the relevant area. In $d=1$ this is simply a stretch of length $\ell$,
            in $d=2$ it should be $\sqrt{\ell^3 r}$,
            hence one should get
            \[ \migrate \sim \rho q(x) s^2 \ell^d \sqrt{r/\ell} (1-e^{-p_e/s}) \]
        \item result does depend on target geometry
    \end{itemize}
}

\reply
As discussed above,
we've debugged the issue about dimensions,
so that the argument now correctly does integrate over the new patch.
The result does depend on target geometry, just as the reviewer points out.
The other part of this sketch
makes further simplifications to \eqref{eqn:migrate}
based on

\point{fluxes}{
}

### 2015-08-30 patchy-selection-paper.tex
.

This suggests that families can be well-described by motion of a central ``trunk'',
which dies at a random time whose distribution is given by $k_e(t)$,
and conditional on survival of this trunk, the family is composed of side ``branches'' off of this trunk,
as depicted in Figure~\ref{fig:branching_decomp}.
We can understand this quite concretely, by a method described rigorously by \citet{geiger1999elementary} (see also \citet{chauvin1991growing}). 
% (lyons1995conceptual deals with conditioned on eventual nonextinction)
When we condition on survival until $t$, we condition on existence of
at least one trunk lineage surviving from time $0$ to time $t$ (the long red
lineage in Figure \ref{fig:branching_decomp}).
Given this lineage, nothing distinguishes any of the other reproduction events --
each individual along the trunk lineage must give birth to at least one offspring that survives until time $t$,
and all other reproduction events are unconditioned.
In other words, the genealogy of a family that survives until $t$
can be decomposed into a trunk that lasts until $t$
and that sprouts independent branches that may or may not survive until $t$.
Specifically, if we write $Z_t$ for the number of offspring in the branching process alive at time $t$,
% (so $\P\{Z_t = 0\} = k_e(t)$),
then this gives the following decomposition:
\begin{align} \label{eqn:Zt_decomposition}
  \left( Z_t \; \vert \; Z_t>0 \right) \deq 1 + \sum_{s=0}^t \sum_{k=1}^{W_s-1} Z^{(s,k)}_{t-s},
\end{align}
where each $Z^{(s,k)}$ is an independent copy of $Z$ 
that arose from the main branch in generation $s$,
and each $W_s$ is independent with distribution $\P\{ W_s = k\} = \P\{ W = k \} (1-k_e(t-s)^k)/(1-k_e(t-s+1))$. \linelabel{ll:branching_process_fix}
Since each subfamily $Z^{(s,k)}$ is a branching process with mean offspring number less than one,
most of these die out quickly,
and so $Z_t$ is composed of a cluster of recently split subfamilies.
If we take the expectation of \eqref{eqn:Zt_decomposition},
since $\E[Z_t] = e^{-s_m t}$,
we get the limiting mean family size conditioned on nonextinction:
\begin{align} \label{eqn:mean_K}
    \E[K] 
    = 1 + \E[W-1 \given W \ge 1] /(1-e^{-s_m}) \approx \E[W-1 \given W \ge 1] / s_m .
\end{align}

@@

Combining these, and absorbing the term $\E[W|W>1]$ from equation \eqref{eqn:mean_K} 
into the constant $C$,
\begin{align} \label{eqn:migrate}
  \migrate(x) &\approx C A \rho s_m \min(s_m,p_e) \left( |y| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |y| \sqrt{2 s_m} / \sigma)  \qquad \text{for } |y| > \sigma/\sqrt{2 s_m}  ,
%     \begin{cases}
%         q(x) \times 4 \rho s_m s_p / \xi^2  \qquad & \text{ if } s_p < s_m \\
%         q(x) \times 2 \rho s_m^2 \qquad & \text{ if } s_p > s_m 
%     \end{cases} .
\end{align}
where $A$ is the area of the new patch if that is no larger than $\sigma/\sqrt{2s_m}$.
When we use this below, we set $p_e = 2 s_p / \xi^2$.
Note that the approximation has the undesirable property that it is
non-monotonic in $s_m$, when we expect the rate of adaptation via
migration to be a decreasing function of $s_m$. However, it only increases for small $s_m$, 
and is a decreasing function of $s_m$ for $s_m > |y|^2 / 2 \sigma^2$, 
which is the parameter region that we are restricted to by other assumptions.
\linelabel{rr:sm_condition}

@@

@@

It is of interest to have an expression for the ``outflux of families'',
$\gamma(r_0)$, 
defined to be the mean proportion of $B$ alleles at distance $r_0$
whose lineage traces back to the original patch without venturing further away than $r_0$.
By expression \eqref{eqn:occupation_integral},
\[
    \gamma(r_0) = \frac{q(y)}{\int_{B(r_0)} u(x,y) dx}
\]
for any $y$.  
Take $\hat x = (1,0)$ and $y=r \hat x$.  Then
\begin{align}
  \frac{ u(x,y) }{ q(y) } &\simeq C' \frac{ \|y-x\|^{-(d-1)/2} e^{-\|y-x\| \sqrt{2s_m}/\sigma } }{ \|y\|^{-(d-1)/2} e^{-\|y\| \sqrt{2s_m}/\sigma } } \\
  &= C' \| \hat x - x/r \|^{-(d-1)/2} e^{-r \sqrt{2s_m}/\sigma  (\|\hat x - x/r\| - 1)} .
\end{align}
Now let $x = (r_0 \cos \theta, r_0 \sin \theta)$,
so that the expression in the exponent is to, first order in $1/r$,
\begin{align}
    r (\|\hat x - x/r\| - 1) &= r \left( \sqrt{ \left( 1 - \frac{r_0}{r} \cos \theta \right)^2 + \frac{r_0^2}{r^2} \sin^2 \theta } - 1 \right ) \\
            &= - r_0 \cos \theta + O(1/r) .
\end{align}
This implies that, still with $r$, $r_0$, and $\theta$ defined as above,
\begin{align}
  \frac{u(x,y)}{q(y)} \longrightarrow_{r \to \infty} C' e^{-r_0 \frac{\sqrt{2s_m}}{\sigma} \cos \theta} .
\end{align}
In one dimension we are done.
In two dimensions,
by bounded convergence and \citet{gradshteyn2007table} (3.339),
\begin{align}
  \int_{B(r_0)} \frac{u(x,y)}{q(y)} dx &\to \gamma(r_0) \quad \text{as } y \to \infty\\
  &= C' \int_0^{2 \pi} e^{-r_0 \frac{\sqrt{2s_m}}{\sigma} \cos \theta} d\theta \\
  &= 2 \pi C' I_0(-r_0\sqrt{2 s_m}/\sigma),
\end{align}
where $I_0$ is a modified Bessel function of order zero of the first kind.

>>>>

This suggests that families can be well-described by motion of a central ``trunk'',
which dies at a random time whose distribution is given by $k_e(t)$,
and conditional on survival of this trunk, the family is composed of side ``branches'' off of this trunk,
as depicted in Figure~\ref{fig:branching_decomp}.
We can understand this quite concretely, by a method described rigorously by \citet{geiger1999elementary} (see also \citet{chauvin1991growing}). 
% (lyons1995conceptual deals with conditioned on eventual nonextinction)
When we condition on survival until $t$, we condition on existence of
at least one trunk lineage surviving from time $0$ to time $t$ (the long red
lineage in Figure \ref{fig:branching_decomp}).
Given this lineage, nothing distinguishes any of the other reproduction events --
each individual along the trunk lineage must give birth to at least one offspring that survives until time $t$,
and all other reproduction events are unconditioned.
In other words, the genealogy of a family that survives until $t$
can be decomposed into a trunk that lasts until $t$
and that sprouts independent branches that may or may not survive until $t$.
Specifically, if we write $Z_t$ for the number of offspring in the branching process alive at time $t$,
% (so $\P\{Z_t = 0\} = k_e(t)$),
then this gives the following decomposition:
\begin{align} \label{eqn:Zt_decomposition}
  \left( Z_t \; \vert \; Z_t>0 \right) \deq 1 + \sum_{s=0}^t \sum_{k=1}^{W_s-1} Z^{(s,k)}_{t-s},
\end{align}
where each $Z^{(s,k)}$ is an independent copy of $Z$ 
that arose from the main branch in generation $s$,
and each $W_s$ is independent with distribution $\P\{ W_s = k\} = \P\{ W = k \} (1-k_e(t-s)^k)/(1-k_e(t-s+1))$, 
where $W$ is the unconditioned offspring number distribution.  \linelabel{ll:branching_process_fix}
Since each subfamily $Z^{(s,k)}$ is a branching process with mean offspring number less than one,
most of these die out quickly,
and so $Z_t$ is composed of a cluster of recently split subfamilies.
If we take the expectation of \eqref{eqn:Zt_decomposition},
since $\E[Z_t] = e^{-s_m t}$,
we get the limiting mean family size conditioned on nonextinction:
\begin{align} \label{eqn:mean_K}
    \E[K] 
    % = 1 + \E[W-1 \given W \ge 1] /(1-e^{-s_m}) 
    = 1 + \sum_{s=0}^\infty \E[W_s-1] e^{-s_m t}
    \approx \E[W-1 \given W \ge 1] / s_m .
\end{align}

@@

Combining these, and absorbing the term $\E[W-1|W>1]$ from equation \eqref{eqn:mean_K} 
into the constant $C$,
\begin{align} \label{eqn:migrate}
  \migrate(x) &\approx C A \rho s_m \min(s_m,p_e) \left( |y| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |y| \sqrt{2 s_m} / \sigma)  \qquad \text{for } |y| > \sigma/\sqrt{2 s_m}  ,
%     \begin{cases}
%         q(x) \times 4 \rho s_m s_p / \xi^2  \qquad & \text{ if } s_p < s_m \\
%         q(x) \times 2 \rho s_m^2 \qquad & \text{ if } s_p > s_m 
%     \end{cases} .
\end{align}
where $A$ is the area of the new patch if that is no larger than $\sigma/\sqrt{2s_m}$.
When we use this below, we set $p_e = 2 s_p / \xi^2$.
Note that the approximation has the undesirable property that it is
non-monotonic in $s_m$, when we expect the rate of adaptation via
migration to be a decreasing function of $s_m$. However, it only increases for small $s_m$, 
and is a decreasing function of $s_m$ for $s_m > |y|^2 / 2 \sigma^2$, 
which is the parameter region that we are restricted to by other assumptions.
\linelabel{rr:sm_condition}

@@

It is of interest to have an expression for the ``outflux of families'',
$\gamma(r_0)$, 
defined to be the mean density of $B$ alleles at distance $r_0$
whose lineage traces back to the original patch without venturing further away than $r_0$.
By expression \eqref{eqn:occupation_integral},
taking $S$ to be a small area located at $y$,
\[
    \gamma(r_0) = \frac{\rho q(y)}{\int_{B(r_0)} u(x,y) dx}
\]
for any $y$.  
Take $\hat x = (1,0)$ and $y=r \hat x$.  Then
\begin{align}
  \frac{ q(y) }{ u(x,y) } &\simeq C' \frac{ \|y\|^{-(d-1)/2} e^{-\|y\| \sqrt{2s_m}/\sigma } }{ \|y-x\|^{-(d-1)/2} e^{-\|y-x\| \sqrt{2s_m}/\sigma } } \\
  &= C' \| \hat x - x/r \|^{(d-1)/2} e^{ r \sqrt{2s_m}/\sigma  (\|\hat x - x/r\| - 1)} .
\end{align}
Now let $x = (r_0 \cos \theta, r_0 \sin \theta)$,
so that the expression in the exponent is to, first order in $1/r$,
\begin{align}
    r (\|\hat x - x/r\| - 1) &= r \left( \sqrt{ \left( 1 - \frac{r_0}{r} \cos \theta \right)^2 + \frac{r_0^2}{r^2} \sin^2 \theta } - 1 \right ) \\
            &= - r_0 \cos \theta + O(1/r) .
\end{align}
This implies that, still with $r$, $r_0$, and $\theta$ defined as above,
\begin{align}
\frac{q(y)}{u(x,y)} \longrightarrow C' e^{-r_0 \frac{\sqrt{2s_m}}{\sigma} \cos \theta} \qquad \text{ as } r \to \infty .
\end{align}
In one dimension we are done (since $\cos \theta=1$).
In two dimensions,
by bounded convergence and \citet{gradshteyn2007table} (3.339),
\begin{align}
  \int_{B(r_0)} \frac{q(y)}{u(x,y)} dx &\to \gamma(r_0) \quad \text{as } y \to \infty\\
  &= C' \int_0^{2 \pi} e^{-r_0 \frac{\sqrt{2s_m}}{\sigma} \cos \theta} d\theta \\
  &= 2 \pi C' I_0(-r_0\sqrt{2 s_m}/\sigma),
\end{align}
where $I_0$ is a modified Bessel function of order zero of the first kind.

@@
